<!--
This is for including Chapter 1.  Notice that it's also good practice to name your chunk.  This will help you debug potential issues as you knit.  The chunk above is called intro and the one below is called chapter1.  Feel free to change the name of the Rmd file as you wish, but don't forget to change it here from chap1.Rmd.
-->

<!--
The {#rmd-basics} text after the chapter declaration will allow us to link throughout the document back to the beginning of Chapter 1.  These labels will automatically be generated (if not specified) by changing the spaces to hyphens and capital letters to lowercase.  Look for the reference to this label at the beginning of Chapter 2.
-->

```{r echo=F}
knitr::opts_chunk$set(fig.width=6, fig.height=4, fig.align='center')
```


<!-- # R Markdown Basics {#rmd-basics} -->

<!-- Here is a brief introduction into using _R Markdown_. _Markdown_ is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. _R Markdown_ provides the flexibility of _Markdown_ with the implementation of **R** input and output.  For more details on using _R Markdown_ see <https://rmarkdown.rstudio.com>.   -->

<!-- Be careful with your spacing in _Markdown_ documents.  While whitespace largely is ignored, it does at times give _Markdown_ signals as to how to proceed.  As a habit, try to keep everything left aligned whenever possible, especially as you type a new paragraph.  In other words, there is no need to indent basic text in the Rmd document (in fact, it might cause your text to do funny things if you do). -->

<!-- ## Lists -->

<!-- It's easy to create a list.  It can be unordered like -->

<!-- * Item 1 -->
<!-- * Item 2 -->

<!-- or it can be ordered like -->

<!-- 1. Item 1 -->
<!-- 4. Item 2 -->

<!-- Notice that I intentionally mislabeled Item 2 as number 4.  _Markdown_ automatically figures this out!  You can put any numbers in the list and it will create the list.  Check it out below. -->

<!-- To create a sublist, just indent the values a bit (at least four spaces or a tab).  (Here's one case where indentation is key!) -->

<!-- 1. Item 1 -->
<!-- 1. Item 2 -->
<!-- 1. Item 3 -->
<!--     - Item 3a -->
<!--     - Item 3b -->

<!-- ## Line breaks -->

<!-- Make sure to add white space between lines if you'd like to start a new paragraph.  Look at what happens below in the outputted document if you don't: -->

<!-- Here is the first sentence.  Here is another sentence.  Here is the last sentence to end the paragraph. -->
<!-- This should be a new paragraph. -->

<!-- *Now for the correct way:*  -->

<!-- Here is the first sentence.  Here is another sentence.  Here is the last sentence to end the paragraph. -->

<!-- This should be a new paragraph. -->

<!-- ## R chunks -->

<!-- When you click the **Knit** button above a document will be generated that includes both content as well as the output of any embedded **R** code chunks within the document. You can embed an **R** code chunk like this (`cars` is a built-in **R** dataset): -->

<!-- ```{r cars} -->
<!-- summary(cars) -->
<!-- ``` -->

<!-- ## Inline code -->

<!-- If you'd like to put the results of your analysis directly into your discussion, add inline code like this: -->

<!-- > The `cos` of $2 \pi$ is `r cos(2*pi)`.  -->

<!-- Another example would be the direct calculation of the standard deviation: -->

<!-- > The standard deviation of `speed` in `cars` is `r sd(cars$speed)`. -->

<!-- One last neat feature is the use of the `ifelse` conditional statement which can be used to output text depending on the result of an **R** calculation: -->

<!-- > `r ifelse(sd(cars$speed) < 6, "The standard deviation is less than 6.", "The standard deviation is equal to or greater than 6.")` -->

<!-- Note the use of `>` here, which signifies a quotation environment that will be indented. -->

<!-- As you see with `$2 \pi$` above, mathematics can be added by surrounding the mathematical text with dollar signs.  More examples of this are in [Mathematics and Science] if you uncomment the code in [Math].   -->

<!-- ## Including plots -->

<!-- You can also embed plots. For example, here is a way to use the base **R** graphics package to produce a plot using the built-in `pressure` dataset: -->

<!-- ```{r pressure, echo=FALSE, cache=TRUE, fig.height=3, fig.width=5} -->
<!-- plot(pressure) -->
<!-- ``` -->

<!-- Note that the `echo=FALSE` parameter was added to the code chunk to prevent printing of the **R** code that generated the plot. There are plenty of other ways to add chunk options (like `fig.height` and `fig.width` in the chunk above).  More information is available at <https://yihui.org/knitr/options/>.   -->

<!-- Another useful chunk option is the setting of `cache=TRUE` as you see here.  If document rendering becomes time consuming due to long computations or plots that are expensive to generate you can use knitr caching to improve performance.  Later in this file, you'll see a way to reference plots created in **R** or external figures. -->

<!-- ## Loading and exploring data -->

<!-- Included in this template is a file called `flights.csv`.  This file includes a subset of the larger dataset of information about all flights that departed from Seattle and Portland in 2014. More information about this dataset and its **R** package is available at <https://github.com/ismayc/pnwflights14>. This subset includes only Portland flights and only rows that were complete with no missing values. Merges were also done with the `airports` and `airlines` data sets in the `pnwflights14` package to get more descriptive airport and airline names. -->

<!-- We can load in this data set using the following commands: -->

<!-- ```{r load_data} -->
<!-- # flights.csv is in the data directory -->
<!-- # string columns will be read in as strings and not factors now -->
<!-- flights <- read.csv('~/tree_imaging/index/data/flights.csv', stringsAsFactors = FALSE) -->
<!-- ``` -->

<!-- The data is now stored in the data frame called `flights` in **R**.  To get a better feel for the variables included in this dataset we can use a variety of functions. Here we can see the dimensions (rows by columns) and also the names of the columns. -->

<!-- ```{r str} -->
<!-- dim(flights) -->
<!-- names(flights) -->
<!-- ``` -->

<!-- Another good idea is to take a look at the dataset in table form.  With this dataset having more than 20,000 rows, we won't explicitly show the results of the command here. I recommend you enter the command into the Console **_after_** you have run the **R** chunks above to load the data into **R**. -->

<!-- ```{r view_flights, eval=FALSE} -->
<!-- View(flights) -->
<!-- ``` -->

<!-- While not required, it is highly recommended you use the `dplyr` package to manipulate and summarize your data set as needed.  It uses a syntax that is easy to understand using chaining operations.  Below I've created a few examples of using `dplyr` to get information about the Portland flights in 2014.  You will also see the use of the `ggplot2` package, which produces beautiful, high-quality academic visuals. -->

<!-- We begin by checking to ensure that needed packages are installed and then we load them into our current working environment: -->

```{r load_pkgs, message=FALSE, warning=F, echo=F}
# List of packages required for this analysis
pkg <- c("dplyr", "ggplot2", "knitr", "bookdown")
# Check if packages are not installed and assign the
# names of the packages not installed to the variable new.pkg
new.pkg <- pkg[!(pkg %in% installed.packages())]
# If there are any packages in the list that aren't installed,
# install them
if (length(new.pkg)) {
  install.packages(new.pkg, repos = "https://cran.rstudio.com")
}
# Load packages
library(thesisdown)
library(tidyverse)
library(ggplot2)
library(knitr)
library(pdxTrees)
library(kableExtra)
```

\clearpage

## Research Problem and Background
Western redcedar trees are evergreen trees that typically grow up to 75 feet tall and are located throughout the Pacific Northwest, making it an organism with a tolerance for shaded regions with moist environments. These trees are native to the land and have served many purposes to people and animals living in the vicinity of the trees, including medicinal, building, and habitat functions [@peterson_western_nodate]. Western redcedars were culturally important to indigenous peoples who valued the strong wood and utilized it for construction and everyday necessities [@noauthor_western_nodate-1]. Over the past decade, reports of dead western redcedars have been increasing, suggesting that something other than natural causes is killing off this species. In general, western redcedars experience several hardships in surviving in the Pacific Northwest, with common causes of death such as forest fires, clearcutting, small animals eating the saplings, and harsh weather including strong winds that easily uproot the trees [@peterson_western_nodate]. Dying redcedar trees can be identified by their branches which turn brownish yellow or fall off completely. Another sign is that the top of a tree will turn brown and lose leaves [@noauthor_western_nodate]. Losing this native tree would have a detrimental effect on animals in the area who rely heavily on the trees for their lifestyle. Scientists have speculated that western redcedar decline might be caused by recent dry summers, the spread of tree disease, insects, or other weather related events [@noauthor_western_nodate]. Since this is a recent issue, there is not a lot of resources explaining the decline. This research aims to provide more insight into the cause of the western redcedar decline by first predicting the location of the western redcedar trees in Portland and then predicting their condition in terms of health. Having more insight helps to prevent further tree deaths and save the western redcedar species, which also extends to similar tree species and provides more knowledge about environmental changes in the Pacific Northwest region.

Modelling in this research will be conducted by combining information gathered from remotely sensed images with ground level information. There are several sources publishing research done using satellite imagery for land classification [@castelluccio_land_2015] and for predicting tree species [@fricker_convolutional_2019], which will be the groundwork for this project's application of remote sensing models to the specific topic of western redcedar mortality. 

Remote sensing is a process for identifying the physical aspects of a large area of land by collecting and measuring the reflected light using airplanes or satellites. This method provides access to lot of information about the region of interest that would otherwise be limited from a human's ground-level persepective. Remote sensing is used to detect characteristics on land as well as the sea. Some remote sensing land applications include tracking forest fires, volcanic eruptions, city growth, and also forest changes.

Satellites orbit the earth, carrying sensors that record levels of electromagnetic radiation detected by the reflection of the sun on the earth. The data collected is the measured radiation from different regions of the spectrum (visible light, infrared, ultraviolet, etc.). Depending on the surface, sunlight gets absorbed or reflected back at the orbiting satellite. The remotely sensed data is available at certain resolutions depending on the instruments used. Spatial resolution is the area imaged by a satellite image, where a detailed image has smaller spatial resolution hence smaller pixels. Spectral resolution is a measure of the size of the wavelength interval, where smaller bands and smaller wavelengths improve the spectral resolution of an image and enhance the level of detail. Satellite images need to be corrected to combine the bands into an image depending on the analysis.

Imagery come in various band widths and different number of bands. Hyperspectral imagery has narrow bands of sizes 10-20nm and hundreds of bands. So, hyperspectral imagery is successful in distinguishing fine details, but at a cost of more complexity. Multispectral imagery has wider bands and up to 10 total bands. This type of imagery is not complex to work with, but misses some details that hyperspectral imagery detects.

Previous literature has been published in utilizing satellite imagery for predictive models, however, many issues arise in classifying land type through remote sensing mainly due to image quality. A single image can be composed of image strips taken over the course of multiple flight paths. Consequently, these images are taken at different times of day, which compiles into a single image with a lot of variation due to changes in the weather as well as the different angles of the sun's position [@castelluccio_land_2015]. Ideally, a solid classifying model surpasses any error introduced by imperfection in the satellite images. Common approaches to classification on satellite images include support vector machines, random forests, and decision trees. One study on land type classification explores the performance of convolutional neural networks (CNN) as classification models [@castelluccio_land_2015].

Another related study identifies 7 tree species by applying a hyperspectral CNN model. This study was completed using field data with measured information about tree count, tree species, and mortality status and hyperspectral imagery data. The hyperspectral data is converted into the form of a tree canopy height model with circular polygons centered at individual tree canopies. The study analyzes the performance of CNNs for both hyperspectral imagery data and a Red-Green-Blue (RGB) subset of the hyperspectral imagery data. The results of the experiment conclude that training a CNN on the hyperspectral data outperforms the CNN trained on RGB data in classifying land type on the UC Merced Land Use dataset. The RGB model does not perform as well as the hyperspectral model in terms of distinguishing tree classes, but does perform around the same level of accuracy in terms of genus classification. For this project, the data comes from Planet.com and only has RGB and IR data available for the Portland region.

This thesis follows along with the methods in [@fricker_convolutional_2019] for combining the satellite imagery data with ground data by creating spatial polygons and extracting pixel-level information to train classification models. The work in this thesis differs from the methods in the cited literature by using random forests and support vector machines as classification models instead of CNNs and uses the RGB model instead of hyperspectral model. Finally, this thesis applies established classification methods for satellite imagery data to answer the specific question about the cause of death of western redcedars in the Pacific Northwest. The models predict the locations for each tree species to facilitate the identification of any patterns in the species mortality over the past decade.

## Overview
### A Statistical Learning Approach to Identifying Location of Western Redcedars
This work combines RGB imagery data from Planet.com with ground level tree data from the RStudio `pdxTrees` library and applies random forest and support vector machine classification methods to predict the location of tree species (specifically western redcedars) in Portland, Oregon and model the condition of western redcedars to ultimately understand the cause of death in this species. Chapter \@ref(data) describes the two levels of data in this research: ground level and pixel level data, where the data comes from, and how the data is combined for modeling. Chapter \@ref(methods) discusses the process behind training random forest and support vector machine models. Chapter \@ref(results) summarizes how the data is prepared for the final model and the outcome of the research.


