---
title: "Final Project"
author: "Sarah Maebius"
date: "5/12/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)
```

Load necessary libraries
```{r}
library(ggplot2)
library(tidyverse)
library(raster)
library(rgdal)
library(tigris)
library(sf)
library(pdxTrees)
library(mecofun)
```

This project explores the distribution of Western Redcedars in Portland. It combines current and predicted climate data from ![WorldClim](https://www.worldclim.org/data/index.html) with tree locations from ![pdxTrees](https://github.com/mcconvil/pdxTrees) to predict and analyse the species distribution changes over time. Methods in this project closely follow methods used in a ![Species Distribution Modelling Study](https://damariszurell.github.io/SDM-Intro/#1_Background). That project investigated the changes in distribution of Ring Ouzel birds in Switzerland using climate data and presence-absence of the species data.

The raster function `getData` was used to obtain a raster of current and predicted climate data for Portland. The variables selected were 

* `bio_2`: the mean of variation between the daily high and low temperatures
* `bio_5`: the maximum temperature of the warmest month
* `bio_14`: the precipitation of the driest month

![Research](https://besjournals.onlinelibrary.wiley.com/doi/10.1111/j.1365-2745.2010.01741.x) suggested that climated change affects tree locations because decreased rainfall and increased temperatures negatively impact tree growth. Some other rainfall and temperature variables were included in the models, but `bio_2`, `bio_5`, and `bio_14` performed the best and they provide measurements for rainfall and temperature.

## Climate analysis
Source: https://www.rdocumentation.org/packages/raster/versions/3.4-5/topics/getData
```{r}
# current climate data
bio_curr <- raster::getData('worldclim', var = 'bio', res = 0.5, lon = -122, 
                    lat = 45, path = 'data')[[c(2, 5, 14)]]
# projected climate data
bio_fut <- raster::getData('CMIP5', var = 'bio', res = 2.5, lon = -122, lat = 45, 
                   rcp = 45, model = 'NO', year = 70, path = 'data', 
                   download = T)[[c(2, 5, 14)]]

# reduce size
pdx_ext <- c(-122.9, -122.4, 45.3, 45.65)
bio_curr <- crop(bio_curr, pdx_ext)
bio_fut <- crop(bio_fut, pdx_ext)

# get portland outline
or_cities <- tigris::places("OR", class = "sp")
portland <- or_cities %>%
  dplyr::filter(NAME == "Portland")
portland_wgs84 <- spTransform(portland, CRS("+proj=utm +zone=10 +ellps=WGS84 
                                            +datum=WGS84 +units=m +no_defs"))

# portland to raster
r <- raster(ncol = 180, nrow = 180)
extent(r) <- extent(portland)
rp <- rasterize(portland, r)

# project and clip to boundary
bio_curr <- projectRaster(bio_curr, rp)
bio_curr <- resample(bio_curr, rp)
bio_curr <- raster::mask(bio_curr, rp)
names(bio_curr) <- c('bio_2', 'bio_5', 'bio_14')
bio_fut <- projectRaster(bio_fut, rp)
bio_fut <- resample(bio_fut, rp)
bio_fut <- mask(bio_fut, rp)
names(bio_fut) <- c('bio_2', 'bio_5', 'bio_14')

# convert temp to celcius
bio_curr[[1]] <- bio_curr[[1]]/10
bio_curr[[2]] <- bio_curr[[2]]/10
bio_fut[[1]] <- bio_fut[[1]]/10
bio_fut[[2]] <- bio_fut[[2]]/10

# visualize climates
plot(bio_curr)
plot(bio_fut)
```

To analyze the climate change over Portland, I computed the rainfall and temperature gradients. The following functions provided code to obtain a semi-systematic sample of coordinates and compute the gradient for a field.

### Functions
```{r}
obtain_sys = function(N,n){
  k = ceiling(N/n)
  r = sample(1:k, 1)
  seq(r, r + k*(n-1), k)
}

GradientComp <- function(field, variable) {
  variable <- field[, variable]
  gradx <- field %>% # x component
  group_by(y) %>%
  mutate(gradx = c(NA, variable[3:n()], NA) - c(NA, variable[1:(n() - 2)], NA))
  grady <- field %>% # y component
    group_by(x) %>%
    mutate(grady = c(NA, variable[3:n()], NA) - c(NA, variable[1:(n() - 2)], NA)) 
  return(full_join(gradx, grady))
}
```

### Current Rainfall and Temperature Gradients
The functions were used to obtain and vizualize the rainfall and temperature gradients for the current climate in Portland.
```{r}
# precipitation
field_cur_precip <- data.frame(rasterToPoints(bio_curr[[3]]))
grad_precip <- GradientComp(field_cur_precip, "bio_14")
gradplot_precip <- grad_precip[obtain_sys(nrow(grad_precip), 100), ]

ggplot(data = field_cur_precip, aes(x, y)) +
  geom_tile(aes(fill = bio_14)) +
  scale_fill_viridis_c() +
  theme_classic() +
  geom_segment(data = gradplot_precip,
               aes(x, y, xend = x + gradx*0.02, yend = y + grady*0.02), 
               arrow = arrow(length = unit(0.01,"npc")),
               color = "maroon") 

# temp range
field_cur_temp <- data.frame(rasterToPoints(bio_curr[[1]]))
grad_temp <- GradientComp(field_cur_temp, "bio_2")
gradplot_temp <- grad_temp[obtain_sys(nrow(grad_temp), 100), ]

ggplot(data = field_cur_temp, aes(x, y)) +
  geom_tile(aes(fill = bio_2)) +
  scale_fill_viridis_c() +
  theme_classic() +
  geom_segment(data = gradplot_temp,
               aes(x, y, xend = x + gradx*0.02, yend = y + grady*0.02), 
               arrow = arrow(length = unit(0.01,"npc")),
               color = "maroon") 

# max temp
field_cur_temp_max <- data.frame(rasterToPoints(bio_curr[[2]]))
grad_temp_max <- GradientComp(field_cur_temp_max, "bio_5")
gradplot_temp_max <- grad_temp_max[obtain_sys(nrow(grad_temp_max), 100), ]

ggplot(data = field_cur_temp_max, aes(x, y)) +
  geom_tile(aes(fill = bio_5)) +
  scale_fill_viridis_c() +
  theme_classic() +
  geom_segment(data = gradplot_temp_max,
               aes(x, y, xend = x + gradx*0.02, yend = y + grady*0.02), 
               arrow = arrow(length = unit(0.01,"npc")),
               color = "maroon") 
```

### Predicted Rainfall and Temperature Gradients
The same was done for the predicted climate variable gradients.
```{r, eval = F}
# precipitation
field_fut_precip <- data.frame(rasterToPoints(bio_fut[[3]]))
grad_precip <- GradientComp(field_fut_precip, "bio_14")

gradx <- field_fut_precip %>% # x component
  group_by(y) %>%
  mutate(gradx = c(NA, bio_14[3:n()], NA) - c(NA, bio_14[1:(n() - 2)], NA))
grady <- field_fut_precip %>% # y component
  group_by(x) %>%
  mutate(grady = c(NA, bio_14[3:n()], NA) - c(NA, bio_14[1:(n() - 2)], NA)) 
grad_precip <- full_join(gradx, grady)

gradplot_precip <- grad_precip[obtain_sys(nrow(grad_precip), 100), ]

ggplot(data = field_fut_precip, aes(x, y)) +
  geom_tile(aes(fill = bio_14)) +
  scale_fill_viridis_c() +
  theme_classic() +
  geom_segment(data = gradplot_precip,
               aes(x, y, xend = x + gradx*0.02, yend = y + grady*0.02), 
               arrow = arrow(length = unit(0.01,"npc")),
               color = "maroon") 

# temp range
field_fut_temp <- data.frame(rasterToPoints(bio_fut[[1]]))
grad_temp <- GradientComp(field_fut_temp, "bio_2")
gradplot_temp <- grad_temp[obtain_sys(nrow(grad_temp), 100), ]

ggplot(data = field_fut_temp, aes(x, y)) +
  geom_tile(aes(fill = bio_2)) +
  scale_fill_viridis_c() +
  theme_classic() +
  geom_segment(data = gradplot_temp,
               aes(x, y, xend = x + gradx*0.02, yend = y + grady*0.02), 
               arrow = arrow(length = unit(0.01,"npc")),
               color = "maroon") 

# max temp
field_fut_temp_max <- data.frame(rasterToPoints(bio_fut[[2]]))
grad_temp_max <- GradientComp(field_fut_temp_max, "bio_5")
gradplot_temp_max <- grad_temp_max[obtain_sys(nrow(grad_temp_max), 100), ]

ggplot(data = field_fut_temp_max, aes(x, y)) +
  geom_tile(aes(fill = bio_5)) +
  scale_fill_viridis_c() +
  theme_classic() +
  geom_segment(data = gradplot_temp_max,
               aes(x, y, xend = x + gradx*0.02, yend = y + grady*0.02), 
               arrow = arrow(length = unit(0.01,"npc")),
               color = "maroon") 
```

## Western Redcedar Data Preparation
Using ![pdxTrees](https://github.com/mcconvil/pdxTrees), I create a table that has Western Redcedar tree presence locations and join that table with climate data for that location. A column `blockCV_tile` is added to separate the data into 4 spatial regions in Portland.

```{r}
# load data
pdxTrees_parks <- get_pdxTrees_parks()
pdxTrees_streets <- get_pdxTrees_streets()
pdxTrees_full <- full_join(pdxTrees_parks, pdxTrees_streets)

pdxTrees <- pdxTrees_full %>%
  mutate(Presence = case_when(
    Common_Name %in% "Western Redcedar" ~ 1,
    TRUE ~ 0
  )) %>%
  dplyr::select(Longitude, Latitude, Presence) %>%
  na.omit()

# combine with climate info
coords <- data.frame(x = pdxTrees$Longitude, y = pdxTrees$Latitude)
points <- SpatialPointsDataFrame(coords, data = pdxTrees, 
                                 proj4string = bio_curr@crs)
values_curr <- extract(bio_curr, points)
values_fut <- extract(bio_fut, points)
trees_curr <- cbind.data.frame(pdxTrees, values_curr) %>% na.omit()
trees_fut <- cbind.data.frame(pdxTrees, values_fut) %>% na.omit()

head(trees_curr)
head(trees_fut)

# spatial blocks
trees_curr <- trees_curr %>%
  mutate(blockCV_tile = case_when(
    Longitude >= min(Longitude) & Longitude < -122.7431 & 
      Latitude <= max(Latitude) & Latitude > 45.57818 ~ 1,
    Longitude >= -122.7431 & Longitude < -122.6531 &
      Latitude <= max(Latitude) & Latitude > 45.57818 ~ 2,
    Longitude >= -122.6531 & Longitude < -122.5631 &
      Latitude <= max(Latitude) & Latitude > 45.57818 ~ 3,
    Longitude >= -122.5631 & Longitude <= max(Longitude) &
      Latitude <= max(Latitude) & Latitude > 45.57818 ~ 4,
    Longitude >= min(Longitude) & Longitude < -122.7431 &
      Latitude <= 45.57818 & Latitude > 45.50606 ~ 5,
    Longitude >= -122.7431 & Longitude < -122.6531 &
      Latitude <= 45.57818 & Latitude > 45.50606 ~ 6,
    Longitude >= -122.6531 & Longitude < -122.5631 &
      Latitude <= 45.57818 & Latitude > 45.50606 ~ 7,
    Longitude >= -122.5631 & Longitude <= max(Longitude) &
      Latitude <= 45.57818 & Latitude > 45.50606 ~ 8,
    Longitude >= min(Longitude) & Longitude < -122.7431 &
      Latitude <= 45.50606 & Latitude >= min(Latitude) ~ 9,
    Longitude >= -122.7431 & Longitude < -122.6531 &
      Latitude <= 45.50606 & Latitude >= min(Latitude) ~ 10,
    Longitude >= -122.6531 & Longitude < -122.5631 &
      Latitude <= 45.50606 & Latitude >= min(Latitude) ~ 11,
    Longitude >= -122.5631 & Longitude <= max(Longitude) &
      Latitude <= 45.50606 & Latitude >= min(Latitude) ~ 12
  ))

ggplot() +
  geom_point(trees_curr, mapping = aes(Longitude, Latitude, color = blockCV_tile))

# separate into train and test data
set.seed(2)
train <- trees_curr %>%
  group_by(Presence) %>%
  slice_sample(prop = 0.7)
test <- anti_join(trees_curr, train)

# compare training data to full data
ggplot(train, aes(Longitude, Latitude, color = Presence)) + 
  geom_point()
ggplot(test, aes(Longitude, Latitude, color = Presence)) + 
  geom_point(alpha = 0.4)
ggplot(trees_curr, aes(Longitude, Latitude, color = Presence)) + 
  geom_point()
```

## Modelling
First a ![generalized linear model](https://cran.r-project.org/web/packages/jtools/vignettes/summ.html) and a random forest is applied to the current climate to predict tree locations.

```{r}
library(jtools)

# Fit GLM
m_glm <- glm(Presence ~ bio_2 + I(bio_2^2) + bio_5 + I(bio_5^2) + bio_14 + I(bio_14^2), family = 'binomial', data = train)

summary(m_glm)
summ(m_glm)
```

```{r, eval = F}
library(ranger)
# Fit RF
# m_rf <- randomForest(Presence ~ ., data = train[, 3:6], importance = T)
# m_rf <- randomForest(y = train[, 3], x = trees_curr[, 4:6], importance = T)
m_rf <- ranger(Presence ~ ., data = train[, 3:6], classification = T)
m_rf
importance(m_rf)
plot(m_rf)
head(getTree(m_rf,1,T))
pred <- c('bio_2', 'bio_5', 'bio_14')
```

## Model Assessment
### Implementing spatial cross validation
```{r, eval = F}
# glm
pred_df <- data.frame(matrix(ncol = 2, nrow = 4))
colnames(pred_df) <- c('accuracy', 'fold')
for(x in 1:12) {
  test_fold <- train %>%
    filter(blockCV_tile != x)
  test_odds <- predict.glm(m_glm, test_fold, type = 'response')
  test_pred <- ifelse(test_odds >= 0.5, 1, 0)
  accuracy <- sum(test_pred == test_fold$Presence)/nrow(test_fold)
  pred_df[x, 'accuracy'] <- accuracy
  pred_df[x, 'fold'] <- x
}

# rf
pred_df <- data.frame(matrix(ncol = 2, nrow = 4))
colnames(pred_df) <- c('accuracy', 'fold')
for(x in 1:12) {
  test_fold <- train %>%
    filter(blockCV_tile != x)
  test_pred <- predict(m_rf, test_fold, type = 'response')
  accuracy <- sum(test_pred$predictions == test_fold$Presence)/nrow(test_fold)
  pred_df[x, 'accuracy'] <- accuracy
  pred_df[x, 'fold'] <- x
}
```

Predictions
```{r, eval = F}
bio_curr_df <- data.frame(rasterToPoints(bio_curr))
bio_curr_df$pred_glm <- ifelse(predict.glm(m_glm, bio_curr_df, type = 'response') >= 0.5, 1, 0)
bio_curr_df$pred_rf <- predict(m_rf, bio_curr_df)$predictions

r_pred_curr <- rasterFromXYZ(bio_curr_df[,-c(3:5)])
plot(r_pred_curr)
plot(rasterFromXYZ(bio_curr_df[, c(1, 2, 6)]))
ggplot(bio_curr_df %>% filter(pred_rf == 1), aes(x, y)) +
  geom_point()
ggplot(bio_curr_df, aes(x, y)) +
  geom_point(alpha = 0.2, fill = "lightgray") +
  geom_point(bio_curr_df %>% filter(pred_rf == 1), mapping = aes(x, y), color = "purple", alpha = 0.8) + theme_minimal()
```

```{r}
bio_fut_df <- data.frame(rasterToPoints(bio_fut))
bio_fut_df$pred_glm <- ifelse(predict.glm(m_glm, bio_fut_df, type = 'response') >= 0.5, 1, 0)
bio_fut_df$pred_rf <- predict(m_rf, bio_fut_df)$predictions

r_pred_fut <- rasterFromXYZ(bio_fut_df[,-c(3:5)])

ggplot(bio_fut_df, aes(x, y)) +
  geom_point(alpha = 0.2, fill = "lightgray") +
  geom_point(bio_fut_df %>% filter(pred_glm == 1), mapping = aes(x, y), color = "purple", alpha = 0.8) + theme_minimal()
```


Questions:
Do redcedars migrate? How long does it take for significant changes in the distribution of redcedars to be visible?
What variables explain the distribution of redcedars?

SDM: https://www.molecularecologist.com/2013/04/23/species-distribution-models-in-r/
Spatial CV: https://geocompr.robinlovelace.net/spatial-cv.html